Collective excitations in competing phases in two and three dimensions

### 2024-03-26 paper/main.tex
method + plots

\section{Iterated equations of motion and Green's functions}\label{sec:ieom}

So far we have discussed the static mean-field equations that essentially describe the static ground state.
These results will be used to compute the quantities, namely expectation values, necessary for the iterated equations of motion approach \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
With it, we can describe two-particle quantities like the various collective excitations of the system.

We start with some operator set $\mathcal{B}$ that is complete with respect to commutation with a Hamiltonian $H$, 
i.e., any commutator $[H, A]$ for any $A \in \mathcal{B}$ can be represented by linear combinations of operators in $\mathcal{B}$.
We can then express any time-dependent operator from this set as
\begin{equation}
    \label{eqn:time_dependent_operator}
    a(t) = \sum_j c_j(t) A_j\,,
\end{equation}
where $c_j(t)$ captures the entire time dependence. 
Inserting this into the Heisenberg equation of motion and applying an operator scalar product $(A_i|\cdot)$ to both sides of the equation yields
\begin{equation}
    \label{eqn:heisenberg}
    \begin{aligned}
        \ddt a(t) = \sum_j \ddt c_j(t) A_j &= \im \sum_j c_j(t) [H, A_j] \\
        \Rightarrow \sum_j \underbrace{(A_i | A_j)}_{\coloneqq \mN_{ij}} \ddt c_j(t) &= \im \sum_j \underbrace{(A_i | [H, A_j])}_{\coloneqq \mM_{ij}} c_j(t) \\
        \Rightarrow \mN \ddt \vec{c}(t) &= \im \mM \vec{c}(t)\,.
    \end{aligned}
\end{equation}
The matrices $\mN$ and $\mM$ contain all the dynamics and energetic properties of the system.
The former is referred to as a dynamical matrix while the latter is called a norm matrix.
The advantage is that one can now handle simple matrices rather than operators acting on an enormous Hilbert space.
The size of these matrices depends entirely on the set of operators.

The matrix $\mM$ is positive semidefinite if the system is in thermal equilibrium.
We prove this statement in the appendix \ref{sec:positive_M}.
We can use this fact to further enhance our ground state phase diagram.
If $\mM$ has even a single negative eigenvalue for any combination of parameters, we know that the system is not in thermal equilibrium, 
i.e., our self-consistency did not capture the correct ground state phase.
This does in fact happen for certain $U<0$ and $V<0$, see the orange shading in \autoref{fig:phase_diagram}.
For finite temperatures, on a square lattice, a phase-separated state as well as the coexistence of a phase-separated state with $s$-wave superconductivity has been found in this region \cite{Linner23}.
The phase boundary that we propose here is qualitatively in good agreement with the one found in the aforementioned reference.
We expect a similar set of phases to be present on the simple cubic lattice but cannot prove this statement at this point.

Additionally, note that while the existence of a negative eigenvalue of $\mM$ proves that the system is not in thermal equilibrium, the absence of one does not prove the contrary.
For example, we do not find any negative eigenvalues of $\mM$ in the $U>0$, $V<0$ region despite previous studies finding $d_{x^2 - y^2}$-wave superconductivity there \cite{Micnas88b,Huang13}.

However, most Hamiltonians do not allow any such operator set to exist as the commutation usually introduces terms of higher order.
For instance, commuting a bilinear term with a quartic Hamiltonian yields, quartic terms. Repeating the commutation yields hexartic ones and so on.
Therefore, one repeats this commutation process and includes the occurring terms within a bigger set,
hoping to get better results as the set size increases.

In this article, we will be using the \textcolor{\markEdited}{symplectic product,}
\begin{equation}
\label{eqn:scalar_product}
    (A | B) \coloneqq  \langle [A^\dagger, B] \rangle\,,
\end{equation}
where the expectation values are taken with respect to the thermal equilibrium of the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}.
Note, that this is not a proper scalar product, as a norm defined using it can also be negative.
For example, assume that for some operator $A$
\begin{equation}
    ||A|| = (A | A) = \langle [A^\dagger, A] \rangle > 0\,,
\end{equation}
then
\begin{equation}
    ||A^\dagger|| = (A^\dagger | A^\dagger) = \langle [A, A^\dagger] \rangle = - ||A|| < 0\,.
\end{equation}
Nevertheless, this does not pose any issue in our calculations.
We will compute $\mM$ by commutating with the full Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} enabling us to capture collective behavior rather than the one particle dynamics given by the mean-field approximation.

\textcolor{\markEdited}{To treat the square and simple cubic lattice on equal footing, we utilize the fact \eqref{eqn:equal_expecs} to create our operator set in $\gamma$-space.}
It consists of operators of the type
\begin{equation}
    \label{eqn:ieom_basis_operator}
    A_\gamma \coloneqq \frac{1}{\sqrt{N}} \sum_{\vk} \delta (\gamma - \hat{\gamma}( \vk )) A_{\vk}\,,
\end{equation}
where $A_{\vk}$ represents each type of operator in \eqref{eqn:operators} and their Hermitian conjugates.
Operators with different indices are put into the set individually, e.g., $n_{\gamma \up}$, and $n_{\gamma \down}$ are distinct operators in our set.
\textcolor{\markEdited}{Additionally, we will include an analogous expression for
\begin{equation}
    \tau_{k} \coloneqq c_{k \uparrow}^\dagger c_{k \downarrow}
\end{equation}
and its Hermitian conjugate.
This will allow us to describe transversal magnons later on.}

Naturally, we have to discretize $\gamma$, 
however, this allows us to obtain good results with drastically smaller matrices compared to a momentum-based operator set and obtain results for a three-dimensional system in the first place.
Practically, we choose $N_\gamma = 6000$ equally spaced discretization points.
A detailed explanation of the numerics is provided in the appendix \ref{sec:numerical_ieom}.
\textcolor{\markEdited}{Computationally, our method is limited by the number of terms included in our operator set as well in the magnitude of $N_\gamma$.}
If the gap $\Delta_\text{tot}$ is of the same order of magnitude as our discretization $\Delta \gamma$ our numerics become inaccurate.
This manifests specifically if a spectral function has strong features at minuscule energies, e.g., the phase mode in the superconducting phase.
In this case, the dynamical matrix $\mM$ picks up negative eigenvalues that vanish if $N_\gamma$ is increased so that $\Delta \gamma$ is sufficiently smaller than $\Delta_\text{tot}$.
This happens if the parameters are chosen from the striped area in \autoref{fig:phase_diagram}.
\textcolor{\markEdited}{This issue only arises in the SC phase, nevertheless, we expect similar inaccuracies in the AFM region for small $U$.
However, we did not perform any computations in that region, hence this does not pose an issue to our analysis.}

Next, we order these operators in a way reminiscent of the $x$- and $p$-terms in the harmonic oscillator, i.e.,
\begin{equation}
    X_i \coloneqq  A_i + A_i^\dagger\,,\quad P_i \coloneqq  A_i - A_i^\dagger\,.
\end{equation}
Then, our set is given by $\mathcal{B}_{XP} \coloneqq \{ X_i, \ldots, P_i, \ldots \}$.
If some operators would be 0 ($P_i$ for any Hermitian term) or duplicates (certain $X_i$ for $g$-type terms, e.g, $g_{k\sigma} = g_{k+Q\sigma}^\dagger$), we omit them.

Due to this choice of set, the matrices occurring in \eqref{eqn:heisenberg} have a special block structure
\begin{equation}
    \label{eqn:xp_set}
    \mM = \begin{pmatrix}
        \mathcal{K}_+ & \kappa \\ \kappa^\dagger & \mathcal{K}_-
    \end{pmatrix}\,,\quad \mN = \begin{pmatrix}
        \Lambda_+ & \mathcal{L} \\ \mathcal{L}^\dagger & \Lambda_-
    \end{pmatrix}\,,
\end{equation}
where the upper block refers to all $X_i$ and the lower block to all $P_i$.
Furthermore, due to symmetry, the relations $\Im [\mathcal{K}_\pm] = \Re [\kappa] = 0$ and $\Re [\Lambda_\pm] = \Im [\mathcal{L}] = 0$ hold.
Thus, if all occurring expectation values are real, both matrices have large empty blocks.
This condition is fulfilled for all cases investigated in this article \textcolor{\markEdited}{because our model has inversion symmetry as well as time-reversal symmetry.}

To study the collective excitations, we investigate various retarded Green's functions 
\begin{equation}
    G_{AB}^\text{ret} (t) = - \im \langle [A(t), B] \rangle \Theta(t)\,,
\end{equation}
and their Fourier-transformed
\begin{equation}
    \label{eqn:standard_gf}
    G_{AB}(z = \omega + \im 0^+) = -\im \int_0^{\infty} e^{\im z t} \langle [A(t), B] \rangle \mathrm{d}t\,,
\end{equation}
where $\Theta(t)$ is the Heaviside function.
Our previous choice of pseudo scalar product enables us to write down a matrix-valued Green's functions in terms of the aforementioned matrices
\begin{align}
    \label{eqn:green_function}
    \mathcal{G}(z &= \omega + \im 0^+) = \mN \frac{1}{-z \mN - \mM} \mN \\
        &\coloneqq  -\mN \mathcal{R}(z) \mN\,,
\end{align}
where we used the resolvent
\begin{equation}
    \label{eqn:resolvent}
    \mathcal{R}(z) \coloneqq  \frac{1}{z \mN + \mM}\,.
\end{equation}
The entries of this matrix $\mathcal{G}(z)$ are the Fourier-transformed Green's functions mentioned in \eqref{eqn:standard_gf} with respect to the operators from the initially chosen set.
\textcolor{\markEdited}{That means, if we index each operator in our set $A_i$, the matrix element $(j,i)$ of $\mathcal{G}$ is the Green's function
\begin{align}
    \mathcal{G}_{ji}(z = \omega +\im 0^+) &=  G_{A_j A_i^\dagger} (z) \nonumber \\
        &= -\im \int_0^{\infty} \langle [A_j(t), A_i^\dagger(0)] \rangle e^{\im z t} \mathrm{d}t\,.
\end{align}}

In the following, we will derive this relation.
\textcolor{\markEdited}{For that, consider the Fourier-transformed Green's function above.
We rewrite both operators in the same manner as in \eqref{eqn:time_dependent_operator} to obtain
\begin{align}
    G_{A_j A_i^\dagger} (z) &= \im \int_0^\infty e^{\im z t} \langle [ A_i^\dagger(0), A_j(t) ] \rangle \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty e^{\im z t} \langle [ \sum_m a_{i,m}^*(0) A_m^\dagger, \sum_n a_{j,n}(t) A_n ] \rangle \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty \sum_{mn} e^{\im z t} a_{i,m}^*(0) \underbrace{\langle [ A_m^\dagger, A_n ] \rangle}_{\equiv \mN_{mn}} a_{j,n}(t) \mathrm{d}t \nonumber \\
        &= \im \int_0^\infty e^{\im z t} \vec{a}_{i}^\dagger (0) \mN \vec{a}_{j}(t) \mathrm{d}t\,.
\end{align}
Furthermore, assuming, that $\mN$ is invertible, we can easily solve the differential equation \eqref{eqn:heisenberg} to obtain
\begin{equation}
    \vec{a}_{j}(t) = \exp \left( \im \mN^{-1} \mM t \right) \vec{a}_{j}(0)\,.
\end{equation}
This allows us to finally obtain
\begin{align}
    \label{eqn:green_derivation}
    G_{A_j A_i^\dagger} (z) &= \im  \int_0^\infty \vec{a}_{i}^\dagger (0) \mN \exp \left( \im \left(\mN^{-1} \mM + z \right) t \right) \vec{a}_{j}(0) \mathrm{d}t  \nonumber \\
        &= - \vec{a}_{i}^\dagger (0) \left[ \mN \frac{1}{\mN^{-1} \mM + z} \right] \vec{a}_{j}(0) \nonumber \\
        &= - \vec{a}_{i}^\dagger (0) \left[ \mN \frac{1}{ \underbrace{\mM + z \mN}}_{\equiv \mathcal{R}(z)} \mN \right] \vec{a}_{j}(0)\,,
\end{align}
where $\vec{a}_j(0)$ and $\vec{a}_i^\dagger (0)$ need to be chosen such, that $A_j = \sum_n a_{j,n} A_n$ and $A_i^\dagger = \sum_n a_{i,n}^* A_n^\dagger$ hold.
The task at hand is finding the resolvent as $\mathcal{R}$.}

To do that, we embrace this matrix structure in \eqref{eqn:xp_set} and assume that all matrix entries are real \textcolor{\markEdited}{so that $\kappa_{ij} = \Lambda_{\pm,ij} = 0$ holds}.
Then, with $\mathcal{R}|_X$ denoting the block of a matrix $\mathcal{R}$ corresponding to the $X$-operators, we can calculate
\begin{align}
    r_X (z) &\coloneqq \mathcal{R}(z)|_X = \left. \frac{1}{\mM + z \mN} \right\vert_X \nonumber \\
        &= \left[ \frac{1}{\mM} - \frac{1}{\mM} z \mN \frac{1}{\mM} + \frac{1}{\mM} z \mN \frac{1}{\mM} z \mN \frac{1}{\mM} - \cdots \right]_X \nonumber \\
        &= \left. \frac{1}{\mM} \sum_{j=0}^\infty \left( -z \mN \frac{1}{\mM} \right)^j \right\vert_X \,.
\end{align}
We may omit every second element of the sum, as we want to obtain the upper left block of the matrix, and $\mN$ always swaps the upper left and lower right block.
Thus, we obtain
\begin{align}
    r_X (z) &= \left. \frac{1}{\mM} \sum_{j=0}^\infty z^{2j} \left( \mN \frac{1}{\mM} \right)^{2j} \right\vert_X \nonumber \\
        &= \left. \frac{1}{\mM - z^2 \mN \mM^{-1} \mN} \right\vert_X \nonumber \\
        &= \frac{1}{\mathcal{K}_+ - z^2 \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger}\,.
\end{align}
Next, we define $\check{N} \coloneqq \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger$ for brevity.
Since $\mathcal{M}$ is positive semidefinite and block diagonal, its blocks $\mathcal{K}_\pm$ must be positive semidefinite as well.
Therefore, $\check{N}$ is also positive semidefinite and we can define its square root.
We obtain
\begin{equation}
    \label{eqn:rx}
    r_X (z) = -\check{N}^{-1/2} \frac{1}{z^2 - \check{N}^{-1/2} \mathcal{K}_+ \check{N}^{-1/2}} \check{N}^{-1/2}\,.
\end{equation}
The matrix $\check{M} \coloneqq \check{N}^{-1/2} \mathcal{K}_+ \check{N}^{-1/2}$ is positive semidefinite and Hermitian by the previous arguments.
If $\check{N}$ has a singular part, the inverse can be replaced by the Moore-Penrose pseudo-inverse.
\textcolor{\markEdited}{A possible way to compute this is to diagonalize $\check{N} = \mathcal{V} \mathcal{D} \mathcal{V}^\dagger$, with a diagonal matrix $\mathcal{D}$ and a unitary transformation matrix $\mathcal{V}$.
Now, the pseudo-inverse can be defined as
\begin{equation}
    \check{N}^+ \coloneqq \mathcal{V} \mathcal{D}^+ \mathcal{V}^\dagger
\end{equation}
with
\begin{equation}
    (\mathcal{D}^+)_{ij} = \begin{cases}
        1/\mathcal{D}_{ij} & \mathcal{D}_{ij} \neq 0 \\ 0 & \text{otherwise}
    \end{cases}\,.
\end{equation}
Note, that this definition yields the standard inverse for invertible matrices.}

The same calculation can be repeated by exchanging $X$ with $P$, yielding the relation for the remaining block.
Then, the definitions of the matrices above change in the following way
\begin{equation}
    \check{N}_P \coloneqq \mathcal{L}^\dagger \mathcal{K}_+^{-1} \mathcal{L}\,,\quad \check{M}_P \coloneqq \check{N}_P^{-1/2} \mathcal{K}_- \check{N}_P^{-1/2}\,.
\end{equation}
In conclusion, the Green's function is given by
\begin{subequations}
    \begin{align}
        G_{AB^\dagger} \vert_P (z) &= - \left[ \vec{b}^\dagger (0) \right]_P \mathcal{L}^\dagger r_X (z) \mathcal{L} \left[ \vec{a}(0) \right]_P\,, \\
        G_{AB^\dagger} \vert_X (z) &= - \left[ \vec{b}^\dagger (0) \right]_X \mathcal{L} r_P (z) \mathcal{L}^\dagger \left[ \vec{a}(0) \right]_X\,,
    \end{align}
\end{subequations}
\textcolor{\markEdited}{where $r_X (z)$ is given in \eqref{eqn:rx} and 
\begin{equation}
    r_P (z) = -\check{N}_P^{-1/2} \frac{1}{z^2 - \check{M}_P} \check{N}_P^{-1/2}\,.
\end{equation}}

>>>>

\section{Iterated equations of motion and Green's functions}
\label{sec:ieom}

So far we discussed the static mean-field equations that describe the ground state.
These results are used to compute the quantities, namely the expectation values, necessary for the iterated equations of motion approach \cite{uhrig09,hamerla13,hamerla14,bleicker18}.
On this basis, we can describe two-particle quantities such as the various collective excitations of the system.

We start with an operator set $\mathcal{B}$ that ideally is complete with respect to commutation with 
the Hamiltonian $H$, i.e., any commutator $[H, A]$ for any $A \in \mathcal{B}$ can be represented by linear combinations of operators in $\mathcal{B}$.
Then, we can express any time-dependent operator in this set by
\begin{equation}
    \label{eqn:time_dependent_operator}
    A(t) = \sum_n a_n(t) A_n,
\end{equation}
where the coefficients $c_j(t)$ capture the entire time dependence. 
Inserting this into the Heisenberg equation of motion and applying a sort of operator scalar product $(A_i|\cdot)$ 
on both sides of the equation yields
\bs
\begin{align}
        \ddt A(t) = \sum_n \ddt a_n(t) A_n &= \im \sum_n a_n(t) [H, A_n] \\
        \Rightarrow \sum_n \underbrace{(A_i | A_n)}_{\coloneqq \mN_{in}} \ddt a_n(t) &= 
				\im \sum_n \underbrace{(A_i | [H, A_n])}_{\coloneqq \mM_{in}} a_n(t) \\
        \Rightarrow \mN \ddt \vec{a}(t) &= \im \mM \vec{a}(t).
				    \label{eqn:heisenberg}
\end{align}
\es
where $\vec{a}:=(a_1, a_2, a_3,\ldots)^\top$.
The matrices $\mN$ and $\mM$ contain all the energetic and dynamic properties of the system.
The former is referred to as norm matrix while the latter is called the dynamic matrix.
The advantage is that one can now handle simple matrices rather than operators 
acting on an enormous Hilbert space.
The dimension of these matrices depends on the number of operators in the set $\mathcal{B}$.

The dynamical matrix $\mM$ is positive semidefinite if the system is in thermal equilibrium.
We derive this statement in  App.\ \ref{sec:positive_M}.
We use this fact to further enhance our ground state phase diagram.
If $\mM$ has even a single negative eigenvalue for a combination of parameters, 
we know that the assumed phase does not represent a true ground state.
Physically, this means that there are ``excitations'' which in fact lower the energy. 
Thus, the assumed ground state is not the lowest state and therefore unstable.
This happens for certain values $U<0$ and $V<0$, see the red shading in Fig.\ \ref{fig:phase_diagram}.
For finite temperatures on the square lattice, a phase-separated state as well as the coexistence of a phase-separated state with $s$-wave superconductivity has been found in this region \cite{Linner23}.
The phase boundary that we propose here agrees qualitatively well with the one found in the aforementioned reference.
We expect a similar set of phases to be present on the simple cubic lattice. \red{This is indicated by
the negative eigenvalues of the corresponding dynamical matrix. Stimmt das so, Joshua?} \blue{ja. Ich überlege gerade, ob dieser Absatz evtl. mehr Sinn nach dem Einführen des symplektischen Produktes hat?}

While the existence of a negative eigenvalue of $\mM$ proves that the assumed phase is
not in thermal equilibrium, the absence of negative eigenvalues does not imply absolute stability,
but only local stability with respect to the considered operators.
For example, we do not find any negative eigenvalues of $\mM$ in the $U>0$, $V<0$ region despite previous studies indicating $d_{x^2 - y^2}$-wave superconductivity \cite{Micnas88b,Huang13}.

In practice, however, generic Hamiltonians do not allow for
a complete operator set $\mathcal{B}$ to exist because the commutations usually introduce terms which are not 
yet in $\mathcal{B}$, e.g., because they are of higher order.
For instance, commuting a bilinear term with a quartic Hamiltonian yields quartic terms. 
Iterating the commutation yields hexatic ones and so on.
One can include such additional terms and generically the results will improve the more
terms are considered in $\mathcal{B}$. In practice, one has to restrict the operators
to a suitable class of terms, thereby truncating certain terms. Here, we will restrict ourselves
to bilinear operators to capture the leading effects of collective behavior.

Furthermore, we use the symplectic product as operator product
\begin{equation}
\label{eqn:scalar_product}
    (A | B) \coloneqq  \langle [A^\dagger, B] \rangle\,,
\end{equation}
where the expectation values are taken with respect to the assumed phase of the mean-field Hamiltonian 
\eqref{eqn:mf_hamiltonian}.
Note that this is not a proper scalar product since it is not positive semidefinite.
For example, let us assume that for some operator $A$
\bs
\begin{equation}
(A | A) = \langle [A^\dagger, A] \rangle > 0,
\end{equation}
then
\begin{equation}
(A^\dagger | A^\dagger) = \langle [A, A^\dagger] \rangle = - (A | A)  < 0.
\end{equation}
\es
This does not pose a serious issue in our calculations but must be kept in mind.
In particular, it implies that the norm matrix is \emph{not positive} as one might have expected.
We emphasize, that the dynamical matrix  $\mM$ is computed
by commuting with the \emph{full} Hubbard Hamiltonian \eqref{eqn:full_hamiltonian} 
enabling us to capture collective behavior 
rather than using only the mean-field Hamiltonian \eqref{eqn:mf_hamiltonian}
which captures the one-particle dynamics, but misses
the two-body dynamics.

In order to treat the square and the simple cubic lattice on equal footing, we exploit 
\eqref{eqn:equal_expecs} to define the operator set in $\gamma$ space
\begin{equation}
    \label{eqn:ieom_basis_operator}
    A_\gamma \coloneqq \frac{1}{\sqrt{N}} \sum_{\vk} \delta (\gamma - \widehat{\gamma}( \vk )) A_{\vk},
\end{equation}
where $A_{\vk}$ represents each type of operator defined in Eqs.~\eqref{eqn:operators} or
their Hermitian conjugates if the operators are not Hermitian themselves.
Operators with different indices are treated separately, e.g., $n_{\gamma \up}$ and $n_{\gamma \down}$ 
are distinct operators in the studied set $\mathcal{B}$.
Additionally, we consider an analogous expression for the operator
\begin{equation}
    \tau_{k} \coloneqq c_{k \uparrow}^\dagger c_{k \downarrow}
\end{equation}
and its Hermitian conjugate.
This will allow us to describe transversal magnons, i.e., perpendicular 
deviations from the staggered magnetization in the AFM phase.

For the numerical treatment, we have to discretize $\gamma$.
Still, this allows us to obtain accurate results with drastically smaller matrices 
compared to an operator set based on a discretization of wave vectors.
This is crucial for dealing with three-dimensional lattices.
Practically, we choose $N_\gamma = 6000$ equally spaced sampling points.
A detailed explanation of the numerics is provided in the App.\ \ref{sec:numerical_ieom}.
Computationally, our method is limited by the number of terms included in our operator set 
as well as by the value of $N_\gamma$.
If the gap $\Delta_\text{tot}$ is of the same order of magnitude as the discretization $t \Delta \gamma$ the numerics become inaccurate.
This manifests specifically if a spectral function has strong features at minute energies, e.g., 
the phase mode in the superconducting phase.
In this case, the dynamical matrix $\mM$ spuriously displays negative eigenvalues that vanish if 
$N_\gamma$ is increased so that $\Delta \gamma$ is sufficiently smaller than $\Delta_\text{tot}$.
This happens if the parameters are chosen from the striped region in Fig.\ \ref{fig:phase_diagram}.
We observe that this issue only arises in the SC phase;
nevertheless, we expect similar inaccuracies in the AFM region for small values of $U$.

In the next step, we rearrange these operators in a way reminiscent of the $x$ and $p$ operators
of a harmonic oscillator
\begin{equation}
    X_i \coloneqq  A_i + A_i^\dagger\,,\quad P_i \coloneqq  A_i - A_i^\dagger.
\end{equation}
Then, the studied operator set is given by $\mathcal{B}_{XP} \coloneqq \{ X_i, \ldots, P_i, \ldots \}$.
If some operators would be 0, e.g., $P_i$ for any Hermitian $A_i$, or duplicate, e.g., certain $X_i$ for $g$-type terms where $g_{k\sigma} = g_{k+Q\sigma}^\dagger$, we omit these redundant operators.

Due to this particular choice of operators, the matrices occurring in \eqref{eqn:heisenberg} acquire a block structure
\begin{equation}
    \label{eqn:xp_set}
    \mM = \begin{pmatrix}
        \mathcal{K}_+ & \kappa \\ \kappa^\dagger & \mathcal{K}_-
    \end{pmatrix}\,,\quad \mN = \begin{pmatrix}
        \Lambda_+ & \mathcal{L} \\ \mathcal{L}^\dagger & \Lambda_-
    \end{pmatrix},
\end{equation}
where the upper block refers to all $X_i$ operators and the lower block to all $P_i$ operators.
Furthermore, due to symmetry, the relations $\Im [\mathcal{K}_\pm] = \Re [\kappa] = 0$ and $\Re [\Lambda_\pm] = \Im [\mathcal{L}] = 0$ hold.
Since the mean-field Hamiltonian is real, all ensuing expectation values are real
and we can conclude both matrices have large empty blocks with zero elements: $\kappa=0$ and $\Lambda_\pm=0$.
This condition is fulfilled for all cases investigated in this article because the model displays
inversion symmetry and time-reversal symmetry \red{or at least the combination of both in the AFM phase}.

To study the collective excitations quantitatively, we investigate various retarded Green's functions 
\bs
\begin{equation}
    G_{AB}^\text{ret} (t) = - \im \langle [A(t), B] \rangle \Theta(t)\,,
\end{equation}
where $\Theta(t)$ is the Heaviside function and their Fourier transforms
\begin{equation}
    \label{eqn:standard_gf}
    G_{AB}(z = \omega + \im 0^+) = -\im \int_0^{\infty} e^{\im z t} \langle [A(t), B] \rangle \mathrm{d}t.
\end{equation}
\es
The choice of the symplectic operator product enables us to write down the
matrix-valued Green's functions in terms of the introduced norm and dynamical matrix
\bs
\begin{align}
    \label{eqn:green_function}
    \mathcal{G}(z &= \omega + \im 0^+) = \mN \frac{1}{-z \mN - \mM} \mN \\
        &\coloneqq  -\mN \mathcal{R}(z) \mN,
\end{align}
where we introduced the resolvent
\begin{equation}
    \label{eqn:resolvent}
    \mathcal{R}(z) \coloneqq  \frac{1}{z \mN + \mM}.
\end{equation}
\es
The entries of this matrix $\mathcal{G}(z)$ are the Fourier-transformed Green's functions given in 
\eqref{eqn:standard_gf} with respect to the studied operators. For simplicity, we formulate the
relation between the Green's functions and the matrices for the operators $A_j$. Eventually, we 
use the block representation resulting for the operators $X_i$ and $P_i$. \red{Wird das klar?}
The matrix element $(j,i)$ refers to
\bs
\begin{align}
    \mathcal{G}_{ji}(z = \omega +\im 0^+) &=  G_{A_j A_i^\dagger} (z) \\
        &= -\im \int_0^{\infty} \langle [A_j(t), A_i^\dagger(0)] \rangle e^{\im z t} \mathrm{d}t.
\end{align}
\es

Performing the  Fourier transformation yields
\bs
\begin{align}
    G_{A_j A_i^\dagger} (z) &= \im \int_0^\infty e^{\im z t} \langle [ A_i^\dagger(0), A_j(t) ] \rangle \mathrm{d}t \
		\\
        &= \im \int_0^\infty e^{\im z t} 
				\langle [ \sum_m a_{i,m}^*(0) A_m^\dagger, \sum_n a_{j,n}(t) A_n ] \rangle \mathrm{d}t 
		\\
        &= \im \int_0^\infty \sum_{mn} e^{\im z t} a_{i,m}^*(0) \underbrace{\langle [ A_m^\dagger, A_n ] \rangle}_{\equiv \mN_{mn}} a_{j,n}(t) \mathrm{d}t 
    \\
        &= \im \int_0^\infty e^{\im z t} \vec{a}_{i}^\dagger (0) \mN \vec{a}_{j}(t) \mathrm{d}t,
\end{align}
\es
where $\vec{a}_j(0)$ and $\vec{a}_i^\dagger (0)$ embody the initial conditions $A_j = \sum_n a_{j,n} A_n$ 
and $A_i^\dagger = \sum_n a_{i,n}^* A_n^\dagger$.
Assuming that $\mN$ is invertible we solve the differential equation \eqref{eqn:heisenberg} 
\begin{equation}
    \vec{a}_{j}(t) = \exp \left( \im \mN^{-1} \mM t \right) \vec{a}_{j}(0).
\end{equation}
Then, the final result reads
\bs
\begin{align}
    \label{eqn:green_derivation}
    G_{A_j A_i^\dagger} (z) &= \im  \int_0^\infty \vec{a}_{i}^\dagger (0) \mN 
		\exp \left( \im \left(\mN^{-1} \mM + z \right) t \right) \vec{a}_{j}(0) \mathrm{d}t 
		\\
        &= - \vec{a}_{i}^\dagger (0) \left[ \mN \frac{1}{\mN^{-1} \mM + z} \right] \vec{a}_{j}(0) 
				\\
        &= - \vec{a}_{i}^\dagger (0) \left[ 
				\mN \mathcal{R}(z) \mN \right] \vec{a}_{j}(0)
\end{align}
\es
with the resolvent $\mathcal{R}(z)$ from \eqref{eqn:resolvent}.
Obviously, the key task is computing the resolvent $\mathcal{R}$ efficiently.

To this end, we exploit the matrix structure in \eqref{eqn:xp_set} and assume that all matrix entries are real so that $\kappa_{ij} = \Lambda_{\pm,ij} = 0$ holds. \red{Stimmt das so?}
Then, with $r_X(z)\coloneqq\mathcal{R}|_{XX}$ denoting the upper left block of a matrix $\mathcal{R}$ which 
refers to the $X$-operators, we can calculate
\bs
\begin{align}
    r_X (z) & = \left. \frac{1}{\mM + z \mN} \right\vert_{XX} \\
        &= \left[ \frac{1}{\mM} - \frac{1}{\mM} z \mN \frac{1}{\mM} + 
				\frac{1}{\mM} z \mN \frac{1}{\mM} z \mN \frac{1}{\mM} - \cdots \right]_{XX}  \\
        &= \left. \frac{1}{\mM} \sum_{j=0}^\infty \left( -z \mN \frac{1}{\mM} \right)^j \right\vert_{XX} .
\end{align}
\es
We can safely omit every second term of the sum since we want to obtain the upper left block of the matrix only
and $\mN$ always swaps between the upper left and lower right block. Thus, we obtain
\bs
\begin{align}
    r_X (z) &= \left. \frac{1}{\mM} \sum_{j=0}^\infty z^{2j} \left( \mN \frac{1}{\mM} \right)^{2j} \right\vert_{XX} 
		\\
        &= \left. \frac{1}{\mM - z^2 \mN \mM^{-1} \mN} \right\vert_{XX} 
				\\
        &= \frac{1}{\mathcal{K}_+ - z^2 \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger}.
\end{align}
\es
For brevity, we define $\check{N}_X \coloneqq \mathcal{L} \mathcal{K}_-^{-1} \mathcal{L}^\dagger$.
Since $\mathcal{M}$ is positive semidefinite and block diagonal its blocks $\mathcal{K}_\pm$ are positive semidefinite as well.
Therefore, $\check{N}_X$ is also positive semidefinite and Hermitian so that its square root can be defined.
Then, the resolvent $r_X(z)$ can be expressed by
\begin{equation}
    \label{eqn:rx}
    r_X (z) = -\check{N}_X^{-1/2} \frac{1}{z^2 - \check{N}_X^{-1/2} \mathcal{K}_+ \check{N}_X^{-1/2}} 
		\check{N}_X^{-1/2},
\end{equation}
where the matrix $\check{M}_X \coloneqq \check{N}_X^{-1/2} \mathcal{K}_+ \check{N}_X^{-1/2}$ is used; it is
positive semidefinite and Hermitian by the previous arguments.

If $\check{N}_X$ has a singular part, for instance, due to numerical inaccuracies \blue{Das kann auch völlig analytisch auftreten}, the inverse can be replaced by the Moore-Penrose pseudo-inverse. 
A possible way to compute this inverse is to diagonalize 
$\check{N}_X = \mathcal{V} \mathcal{D} \mathcal{V}^\dagger$ yielding a diagonal matrix $\mathcal{D}$ and a unitary transformation matrix $\mathcal{V}$. 
Then, the pseudo-inverse can be defined by
\begin{equation}
    \check{N}^{-1} \coloneqq \mathcal{V} \mathcal{D}^{-1} \mathcal{V}^\dagger
\end{equation}
with \red{Ich habe das hoch + in ein hoch -1 geaendert, weil mir das "`+"' zu sehr an Transposition erinnerte.}
\begin{equation}
    (\mathcal{D}^{-1})_{ij} = \begin{cases}
        1/\mathcal{D}_{ij} & \mathcal{D}_{ij} \neq 0 \\ 0 & \text{otherwise}
    \end{cases}.
\end{equation}
Note that this definition recovers the standard inverse for invertible matrices.

The same calculation can be repeated by exchanging $X$ with $P$, yielding the relation for the lower right block.
The relevant matrices turn out to be 
\bs
\begin{align}
    \check{N}_P &\coloneqq \mathcal{L}^\dagger \mathcal{K}_+^{-1} \mathcal{L}
		\\
		\check{M}_P &\coloneqq \check{N}_P^{-1/2} \mathcal{K}_- \check{N}_P^{-1/2}\,.
\end{align}
\es
Summarizing, the Green's function is given by
\begin{subequations}
    \begin{align}
        G_{X_j,X_i}  (z) &= - \left[ \vec{x}_j^\dagger (0) \right]_X \mathcal{L}^\dagger r_X (z) \mathcal{L} \left[ \vec{x}_i(0) \right]_X, 
				\\
        G_{P_j,P_i} (z) &= \red{+} \left[ \vec{p}_j^\dagger (0) \right]_P \mathcal{L} r_P (z) \mathcal{L}^\dagger \left[ \vec{p}_i(0) \right]_P\,,
    \end{align}
\end{subequations}
\blue{Brauchen/wollen wir die $[]_{X/P}$ noch? Und ich bin mir ziemlich sicher, dass das in beiden Fällen ein Minus sein sollte.}
where the vectors $x_j(0)$ and $x_i(0)$ describe the initial conditions for the operators
$X_j$ and $X_i$, respectively and $p_j(0)$ and $p_i(0)$ correspondingly for the operators
$P_j$ and $P_i$. The resolvent $r_X (z)$ is given in \eqref{eqn:rx} and 
\begin{equation}
    r_P (z) = -\check{N}_P^{-1/2} \frac{1}{z^2 - \check{M}_P} \check{N}_P^{-1/2} .
\end{equation}
\red{@ Joshua: Bitte noch einmal sorgfaeltig alles pruefen, da ich doch einiges angepasst habe.}
